<span style='line-height:60px; font-size:60px'>Changing communicative need<br>predicts lexical competition<br>and contributes<br>to language change</span></font>

### 2019-08-19 ruse2019/index.Rmd
ruse talk files

---
title: "<span style='line-height:60px; font-size:80px'>Catching competitors</span><font size=12>:</font><br><font size=8><span style='line-height:60px'>lexical evolution in diachronic corpora</span></font>"

author: "<font size=6>Andres Karjus, Richard A. Blythe, Simon Kirby, Kenny Smith<br></font><font size=5>Centre for Language Evolution, University of Edinburgh</font>"
date: "<font size=5>CL2019, Cardiff, July 2019</font>"
output:
  xaringan::moon_reader:
    css: img/xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
---
class: inverse
# All languages change

@@

```{r}
library(plotly, quietly = T)
load("/Users/pplsuser/Dropbox/phd/lexical_competition/lexcom_poster/posterplots.RData")
load("/Users/pplsuser/Dropbox/phd/lexical_competition/isle/dat2.RData")
load("/Users/pplsuser/Desktop/korpused/lexcom_model_twt/twtplots.RData")
```

@@

```{r, out.height=440, out.width="80%", dpi=500, eval=T}
#<font size="3pt">
library(magrittr)
library(plotly, quietly = T)
trendplot %>% layout(legend = list(orientation = 'h', font = list(size = 12)))
#plot(1)
```

@@

```{r, out.width="70%", fig.height=7, eval=T}
expl
```

@@

```{r, out.width='80%', fig.height=7, eval=T}
library(magrittr)
library(plotly, quietly = T)
  # tags$div(resultsplot, style = 
  #   "width: 100%; margin-top:0em !important; margin-bottom:0em !important; border: solid;")
resultsplot %>%
  add_annotations(
    x= -0.1,
    y= 0.52,
    xanchor="left",
    xref = "x",
    yref = "y",
    text = "English (COHA, 1890-1990)",
    showarrow = F
  )
#%>% layout(margin=list( l = 40, r = 1, b = 0, t = 0, pad = 0 ))
#par(mar=c(0,0,0,0))
#plot(1)
```

@@

- Real effect is bigger?
    - Just words; messy population aggregates; messy ML
    - But also, this approach relies on a *lot* of parameters
--
- How does this relate to individual utterance selection processes?

---
# Work in progress: Scottish Twitter

- ~50k daily users; 72 days, ~1.5m words/day, >140k unique #s
```{r, out.width='60%',fig.height=4, dpi=300}
library(ggplot2)
gotplot
```

---
# Results (Scottish Twitter hashtags)

.small[ R<sup>2</sup>=0.12, advection p=0.34 ]
```{r, out.width='90%', fig.height=7, eval=T}
resultsplot_twt
```

---
class:inverse
background-image: url(img/dany.png) 
background-size: cover

@@

- Small set of targets (63)? Also tried with 101 (any) words.
- What do #s compete with anyway: only other #s, or all words?
- Timespans need more thought.
- Probably behave differently from "normal" words:
- &nbsp;&nbsp;&nbsp; *I watched Game of Thrones last night.*
--

@@

- Communicative need describes a small amount of variance in competitive interactions between words in diachronic corpora (but not Twitter).
- Presumably high communicative need facilitates the co-existence of similar words.
--

- Future directions: explore parameters; test with phrases; more corpora; implement with non-discrete spans and modern semantics models .small[ (e.g. temporal referencing, Dubossarsky et al 2019) ]; test this experimentally.
--

- Slides: andreskarjus.github.io/cl2019talk <br>Twitter: @AndresKarjus

@@

- preprocessing choices (lemmatized; removed stopwords, numbers; homogenized compounds, spelling) <br>
- timespan (10y), min change (log()>2), min frequency in t2 (100), min occurrence years (2); twitter: 5d spans, at least once in span >=0.5% daily users <br>
- LSA k (100 dim), min freq (100), context window size (5), weighted <br>
- cosine distance: normalized by 1st neighbour <br>
- density: LSA-based, as mean of cos.sim of 2nd..10th neighbours <br>
- form similarity (restricted Damerau-Levenshtein, length-normalized) <br>
- filter out: leftover>100%, polysemy residuals>2 <br>
- advection topic model k (75 words), min freq (100), context window size (10), weighted <br>
- polysemy model, context window size (2), weighted

@@

```{r, fig.height=5, out.width="50%"}
displot
```

>>>>

---
title: "<span style='line-height:60px; font-size:60px'>Changing communicative need<br>predicts lexical competition<br>and contributes<br>to language change</span></font>"
author: "<font size=6>Andres Karjus, Richard A. Blythe, Simon Kirby, Kenny Smith<br></font><font size=5>Centre for Language Evolution, University of Edinburgh</font>"
date: "<font size=5>RUSE Symposium, Manchester, August 2019</font>"
output:
  xaringan::moon_reader:
    css: img/xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
      ratio: "16:9"
      navigation:
        scroll: false
---
class: inverse
# All languages change

@@

```{r}
#library(plotly, quietly = T)
library(magrittr)
load("/Users/s1364178/Dropbox/phd/lexical_competition/lexcom_poster/posterplots.RData")
load("/Users/s1364178/Dropbox/phd/lexical_competition/isle/dat2.RData")
#load("/Users/s1364178/Desktop/korpused/lexcom_model_twt/twtplots.RData")
```

@@

```{r, out.height=440, out.width="80%", dpi=500, eval=F}
#<font size="3pt">
library(magrittr)
library(plotly, quietly = T)
trendplot %>% layout(legend = list(orientation = 'h', font = list(size = 12)))
#plot(1)
```
<img src="img/trends.png" width="80%" />

@@

```{r examplespace, out.width="95%", fig.height=3, dpi=500, eval=T}
#expl
library(ggplot2)
ex = data.frame(x={set.seed(1);rnorm(1000)}, y={set.seed(2);rnorm(1000)},change=rnorm(1000,sd=2.4) );ex$change[1]=10;ex$change[2]=-10; ex$change[200:1000] = rnorm(801, sd=0.2)
rownames(ex)=1:1000
cowplot::plot_grid(
ggplot(ex, aes(x,y, color=change)) + 
  annotate("line", x=ex[c(446, 656, 963),"x"], y=ex[c(446, 656, 963),"y"], color="gray") +
  geom_point(size=1.4, alpha=0.7) +
  geom_point(aes(x=ex[446,"x"], y=ex[446,"y"]), color=rgb(1,0,0), size=2.2) +
  geom_point(aes(x=ex[656,"x"], y=ex[656,"y"]), color="#2B2BF0", size=1.5) +
  geom_point(aes(x=ex[963,"x"], y=ex[963,"y"]), color="#8B8BCF", size=1.5)+
  theme_minimal(base_size = 10) + 
    theme(legend.position = "right") +
    scale_color_gradient2(low="blue", mid="lightgray", high="red") +
  labs(x="", y="") + theme(legend.position = "none")
,

ggplot(ex, aes(x,y, color=change)) + 
  annotate("line", x=ex[c(446, 656, 963),"x"], y=ex[c(446, 656, 963),"y"], color="black", size=1)+
  geom_point(size=4.4, alpha=0.7) +
  geom_point(aes(x=ex[446,"x"], y=ex[446,"y"]), color=rgb(1,0,0), size=9) +
  geom_point(aes(x=ex[656,"x"], y=ex[656,"y"]), color="#2B2BF0", size=4.6) +
  geom_point(aes(x=ex[963,"x"], y=ex[963,"y"]), color="#8B8BCF", size=4.6)+
  annotate("text", x=ex[c(446, 656, 963),"x"]+0.13, y=ex[c(446, 656, 963),"y"], label=c("+10", "-8", "-2"), color="black", size=3) +
  theme_minimal(base_size = 10) + 
    theme(legend.position = "right") +
    scale_color_gradient2(low="blue", mid="lightgray", high="red", name= "change\npmw") +
  labs(x="", y="") + lims(x=c(-3.2,-2.2), y=c(-0.5,0.4))
)

@@

- R<sup>2</sup>=0.2

```{r, out.width='80%', fig.height=7, eval=F}
library(magrittr)
library(plotly, quietly = T)
  # tags$div(resultsplot, style = 
  #   "width: 100%; margin-top:0em !important; margin-bottom:0em !important; border: solid;")
resultsplot %>%
  add_annotations(
    x= -0.1,
    y= 0.52,
    xanchor="left",
    xref = "x",
    yref = "y",
    text = "English (COHA, 1890-1990)",
    showarrow = F
  )
#%>% layout(margin=list( l = 40, r = 1, b = 0, t = 0, pad = 0 ))
#par(mar=c(0,0,0,0))
#plot(1)

<img src="img/resultsplot.png" width="90%"/>

@@

- Real effect could be bigger?
    - Just words; messy population aggregates; messy ML
    - But also, this approach relies on a *lot* of parameters
--
- How does this relate to individual utterance selection processes?

---
background-image: url(img/tweets.png) 
background-size: cover

# Work in progress: Scottish Twitter

- ~50k daily users; 74 days<br> ~2.7m words/day (incl. retweets)

---
background-image: url(img/tweets.png) 
background-size: cover

# Work in progress: Scottish Twitter

- ~50k daily users; 74 days<br> ~2.7m words/day (incl. retweets)

<img src="img/twplot.png" width="62%" />

---
# Results (Scottish Twitter)

- R<sup>2</sup>=0.29
```{r, out.width='90%', fig.height=7, eval=T}
#resultsplot_twt
```

<img src="img/twt_results.png" width="85%" />
---
# Results (Scottish Twitter)

- R<sup>2</sup>=0.29. Clearer competition signal if: 
    - lower communicative need (advection; $\beta = 0.06$, $p<0.001$)
    - lower frequency change
    - bursty series
    - a clear loser present
    - frequency change supported by retweets*
- (plus controlled again for all the other lexicostatistical variables)
---
class:inverse
background-image: url(img/dany.png) 
background-size: cover

@@

- *No advection effect when: no lemmatization & no retweets
- How to treat retweets & likes (here: multiplied by rt, targets filtered to be >10% from original)
- Timespans need more thought (here: 5 days)
- Pre-processing questions (#, @)
- Probably different from "normal" communication:
- &nbsp;&nbsp;&nbsp; *I watched Game of Thrones last night.*
--

@@

- Communicative need describes a small amount of variance in competitive interactions between words in diachronic corpora
- Presumably high communicative need facilitates the co-existence of similar words.
--

- Future directions: explore parameters; test with phrases; more corpora; implement with non-discrete spans, modern semantics models .small[ (e.g. temporal referencing, Dubossarsky et al 2019) ]; test this experimentally.
--

- Slides: andreskarjus.github.io/ruse2019 <br>Twitter: @AndresKarjus <br>Also: _Drift, selection, change, and shift_, workshop at Evolang2020!

@@

- preprocessing choices (lemmatized; removed stopwords, numbers; homogenized compounds, spelling) <br>
- timespan (10y), min change (log()>2), min frequency in t2 (100), min occurrence years (2); Twitter: 5d spans <br>
- LSA k (100 dim), min freq (100), context window size (5), weighted <br>
- cosine distance: normalized by 1st neighbour <br>
- density: LSA-based, as mean of cos.sim of 2nd..10th neighbours <br>
- form similarity (restricted Damerau-Levenshtein, length-normalized) <br>
- filter out: leftover>100%, polysemy residuals>2 <br>
- advection topic model k (75 words), min freq (100), context window size (10), weighted <br>
- polysemy model, context window size (2), weighted <br>
- Twitter further constraints: 500 min freq in t2, min 50 on first day of t2, min 10% freq must be from original (not re)tweets

@@

```{r, fig.height=5, out.width="50%", eval=F}
displot
```
<img src="img/polysemy.png" width="40%"/>
